---
title: "Determinants of Household Electricity Consumption"
subtitle: "Statistical Analysis in R"
author: "Yiwu Ongum Ihimbru"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

# Create figures directory if it does not already exist
dir.create("../figures", showWarnings = FALSE)
```

## Project Context

This analysis was developed in the context of the **Project Learning from Data** course at Hasselt University.

The workflow includes data validation, multiple linear regression, model selection, diagnostic assessment, standardised coefficient comparison, and interaction modelling.

> **Data availability:** The original academic dataset is not publicly distributed with this repository. To reproduce the analysis locally, place the dataset in the `data/` folder as `data_students.txt`.

---

## 1. Load Libraries

```{r libraries}
library(tidyverse)
library(car)
library(MASS)
library(knitr)
```

---

## 2. Load Data and Apply Group-Specific Row Removal

```{r load-data}

# Relative path used so that the analysis is portable across computers
data_raw <- read.csv(
  "../data/data_students.txt",
  sep = ";",
  header = TRUE,
  stringsAsFactors = FALSE
)

cat(
  "Raw dimensions:",
  nrow(data_raw), "rows x",
  ncol(data_raw), "columns\n"
)

# Group-specific row removal rules (g = 7):
# Row 49  = g^2       = 7^2       = 49
# Row 77  = g * 11    = 7 * 11    = 77
# Row 154 = 19g + 21  = 19(7) + 21 = 154

rows_to_remove <- c(49, 77, 154)

data <- data_raw[-rows_to_remove, ]

cat(
  "Rows after group-specific removal:",
  nrow(data), "\n"
)
```

---

## 3. Variable Coding

```{r variable-coding}

# EPC label: reference category = A (most efficient)
data$epc_label <- factor(
  data$epc_label,
  levels = c("A", "B", "C", "D", "E", "F")
)

# Building era: reference category = Post-2000
data$building_era <- factor(
  data$building_era,
  levels = c("Post-2000", "1970-2000", "Pre-1970")
)

# Roof colour: reference category = Black
data$roof_color <- factor(
  data$roof_color,
  levels = c("Black", "Red", "Other")
)

# Pet ownership: binary factor
data$pet_ownership <- factor(data$pet_ownership)

str(data)
```

---

## 4. Missing Data

```{r missing-data}

# Count and percentage of missing observations per variable
miss_count <- colSums(is.na(data))
miss_pct <- round(colMeans(is.na(data)) * 100, 2)

miss_summary <- data.frame(
  Variable = names(miss_count),
  N_missing = miss_count,
  Pct_missing = miss_pct
)

# Display variables containing missing observations
miss_summary[miss_summary$N_missing > 0, ]

# Missingness was below the prespecified 10% threshold.
# Complete-case analysis was therefore applied.
data_cc <- data[complete.cases(data), ]

cat(
  "Complete cases:",
  nrow(data_cc),
  "out of",
  nrow(data),
  "\n"
)
```

---

## 5. Data Validation

```{r data-validation}

# Inspect annual electricity consumption before modelling
summary(data_cc$annual_kwh)

# Five observations contained implausibly large electricity-use values,
# ranging from approximately 2.9 to 5.4 million kWh.
#
# The highest legitimate value was 16,395 kWh, producing a very large
# natural gap in the outcome distribution.
#
# A conservative threshold of 100,000 kWh therefore identifies the
# same five apparent data-entry errors.

implausible <- data_cc[
  data_cc$annual_kwh > 100000,
]

cat(
  "Implausible observations identified:",
  nrow(implausible),
  "\n"
)

print(
  implausible[
    ,
    c(
      "annual_kwh",
      "epc_label",
      "sq_meters",
      "occupancy_count"
    )
  ]
)

# Remove apparent recording errors before model fitting
data_clean <- data_cc[
  data_cc$annual_kwh <= 100000,
]

cat(
  "Final analytic sample:",
  nrow(data_clean),
  "observations\n"
)

cat(
  "Maximum annual_kwh after validation:",
  max(data_clean$annual_kwh),
  "\n"
)
```

---

## 6. Descriptive Statistics

### Continuous Variables

```{r descriptives-continuous}

cont_vars <- c(
  "annual_kwh",
  "sq_meters",
  "income_euro",
  "occupancy_count",
  "dist_to_brussels"
)

summary(data_clean[, cont_vars])
```

### Categorical Variables

```{r descriptives-categorical}

for (
  var in c(
    "epc_label",
    "building_era",
    "roof_color",
    "pet_ownership"
  )
) {

  cat("\n---", var, "---\n")

  tbl <- table(data_clean[[var]])

  print(
    cbind(
      N = tbl,
      Pct = round(
        100 * prop.table(tbl),
        1
      )
    )
  )
}
```

---

## 7. Full Multiple Linear Regression Model

```{r full-model-clean}

model_full_clean <- lm(
  annual_kwh ~
    sq_meters +
    epc_label +
    building_era +
    income_euro +
    occupancy_count +
    dist_to_brussels +
    roof_color +
    pet_ownership,
  data = data_clean
)

summary(model_full_clean)
```

---

## 8. Diagnostics for the Full Model

```{r diagnostics-clean, fig.width=8, fig.height=6}

par(mfrow = c(2, 2))

plot(
  model_full_clean,
  pch = 16,
  cex = 0.5
)

par(mfrow = c(1, 1))
```

### Residual Distribution

```{r histogram-clean}

hist(
  residuals(model_full_clean),
  main = "Histogram of residuals - clean full model",
  xlab = "Residuals"
)
```

### Q-Q Plot

```{r qqplot-clean}

qqnorm(
  residuals(model_full_clean),
  main = "Q-Q plot - clean full model"
)

qqline(
  residuals(model_full_clean)
)
```

### Transformation Decision

```{r assumption-decision}

cat("Outcome: annual_kwh on original scale\n")
cat("Transformation: none\n")
cat(
  "Decision: OLS assumptions were sufficiently satisfied after data validation.\n"
)
```

---

## 9. Model Selection: Research Question 1

Bidirectional stepwise AIC was applied starting from the full model.

```{r stepwise-aic}

model_selected <- MASS::stepAIC(
  model_full_clean,
  direction = "both",
  trace = TRUE
)
```

### Preferred Model

```{r selected-summary}

summary(model_selected)

cat(
  "\nAIC of preferred model:",
  round(AIC(model_selected), 2),
  "\n"
)

confint(model_selected)
```

---

## 10. Incremental Model Comparison

```{r block-models}

# Block 1: Structural predictors
model_b1 <- lm(
  annual_kwh ~
    sq_meters +
    epc_label +
    building_era,
  data = data_clean
)

# Block 2: Add socioeconomic predictors
model_b2 <- lm(
  annual_kwh ~
    sq_meters +
    epc_label +
    building_era +
    income_euro +
    occupancy_count,
  data = data_clean
)

# Block 3: Add geographic predictor
model_b3 <- lm(
  annual_kwh ~
    sq_meters +
    epc_label +
    building_era +
    income_euro +
    occupancy_count +
    dist_to_brussels,
  data = data_clean
)

# Block 4: Add screening predictors
model_b4 <- lm(
  annual_kwh ~
    sq_meters +
    epc_label +
    building_era +
    income_euro +
    occupancy_count +
    dist_to_brussels +
    roof_color +
    pet_ownership,
  data = data_clean
)

model_comparison <- data.frame(

  Model = c(
    "Block 1 - Structural",
    "Block 2 - Socioeconomic",
    "Block 3 - Geographic",
    "Block 4 - Screening",
    "Preferred - Stepwise AIC"
  ),

  Adj_R2 = round(
    c(
      summary(model_b1)$adj.r.squared,
      summary(model_b2)$adj.r.squared,
      summary(model_b3)$adj.r.squared,
      summary(model_b4)$adj.r.squared,
      summary(model_selected)$adj.r.squared
    ),
    3
  ),

  AIC = round(
    c(
      AIC(model_b1),
      AIC(model_b2),
      AIC(model_b3),
      AIC(model_b4),
      AIC(model_selected)
    ),
    1
  )
)

kable(
  model_comparison,
  caption = "Model comparison across predictor blocks"
)
```

---

## 11. Preferred Model Diagnostics

```{r diagnostics-selected, fig.width=8, fig.height=6}

par(mfrow = c(2, 2))

plot(
  model_selected,
  pch = 16,
  cex = 0.5
)

par(mfrow = c(1, 1))
```

### Multicollinearity

```{r vif-selected}

# Generalised Variance Inflation Factor
# Threshold:
# GVIF^(1/(2*df)) < sqrt(5) ≈ 2.24

vif_vals <- car::vif(model_selected)

print(
  round(vif_vals, 3)
)

vif_check <- if (
  is.matrix(vif_vals)
) {
  vif_vals[, 3]
} else {
  vif_vals
}

cat(
  "\nAll adjusted GVIF values below 2.24:",
  all(vif_check < sqrt(5)),
  "\n"
)
```

### Leverage

```{r leverage-selected}

# Threshold:
# h_ii > 2(p+1)/n

p <- length(coef(model_selected)) - 1

leverage_threshold <-
  2 * (p + 1) / nrow(data_clean)

hat_values <-
  hatvalues(model_selected)

n_leverage <-
  sum(hat_values > leverage_threshold)

cat(
  "Leverage threshold:",
  round(leverage_threshold, 4),
  "\n"
)

cat(
  "High-leverage observations:",
  n_leverage,
  "\n"
)

cat(
  "Percentage of sample:",
  round(
    100 * n_leverage / nrow(data_clean),
    1
  ),
  "%\n"
)
```

---

## 12. Save Preferred Model Diagnostics

```{r save-diag-preferred}

png(
  "../figures/diagnostics_preferred.png",
  width = 800,
  height = 600
)

par(mfrow = c(2, 2))

plot(
  model_selected,
  pch = 16,
  cex = 0.5
)

par(mfrow = c(1, 1))

dev.off()

cat(
  "Saved: figures/diagnostics_preferred.png\n"
)
```

---

## 13. Research Question 2: EPC Label Effects

```{r rq2-epc}

# EPC coefficients relative to reference category A

epc_rows <- grep(
  "epc_label",
  names(coef(model_selected))
)

epc_coefs <- data.frame(

  Label = gsub(
    "epc_label",
    "Label ",
    names(coef(model_selected))[epc_rows]
  ),

  Estimate = round(
    coef(model_selected)[epc_rows],
    1
  ),

  CI_lower = round(
    confint(model_selected)[epc_rows, 1],
    1
  ),

  CI_upper = round(
    confint(model_selected)[epc_rows, 2],
    1
  ),

  p_value = round(
    summary(model_selected)$coefficients[
      epc_rows,
      4
    ],
    4
  )
)

rownames(epc_coefs) <- NULL

kable(
  epc_coefs,
  caption =
    "EPC label coefficients relative to label A (kWh)"
)
```

---

## 14. Research Question 3: Household Income

```{r rq3-income}

income_coef <-
  summary(model_selected)$coefficients[
    "income_euro",
  ]

cat(
  "Income coefficient:",
  round(income_coef["Estimate"], 6),
  "\n"
)

cat(
  "p-value:",
  round(income_coef["Pr(>|t|)"], 3),
  "\n"
)

cat(
  "95% CI: [",
  round(
    confint(model_selected)[
      "income_euro",
      1
    ],
    6
  ),
  ",",
  round(
    confint(model_selected)[
      "income_euro",
      2
    ],
    6
  ),
  "]\n"
)
```

### Standardised Coefficients

```{r rq3-standardised}

sds <- sapply(
  data_clean[
    ,
    c(
      "annual_kwh",
      "sq_meters",
      "income_euro",
      "occupancy_count",
      "dist_to_brussels"
    )
  ],
  sd,
  na.rm = TRUE
)

coefs <- coef(model_selected)

std_coefs <- c(

  sq_meters =
    unname(coefs["sq_meters"]) *
    sds["sq_meters"] /
    sds["annual_kwh"],

  income_euro =
    unname(coefs["income_euro"]) *
    sds["income_euro"] /
    sds["annual_kwh"],

  occupancy_count =
    unname(coefs["occupancy_count"]) *
    sds["occupancy_count"] /
    sds["annual_kwh"],

  dist_to_brussels =
    unname(coefs["dist_to_brussels"]) *
    sds["dist_to_brussels"] /
    sds["annual_kwh"]
)

cat(
  "\nStandardised coefficients for continuous predictors:\n"
)

print(
  round(
    sort(
      abs(std_coefs),
      decreasing = TRUE
    ),
    3
  )
)
```

### Correlation Analysis

```{r rq3-correlation}

cor_income_kwh <- cor(
  data_clean$income_euro,
  data_clean$annual_kwh,
  use = "complete.obs"
)

cor_income_sqm <- cor(
  data_clean$income_euro,
  data_clean$sq_meters,
  use = "complete.obs"
)

cat(
  "Correlation: income vs annual_kwh:",
  round(cor_income_kwh, 3),
  "\n"
)

cat(
  "Correlation: income vs sq_meters:",
  round(cor_income_sqm, 3),
  "\n"
)
```

---

## 15. Research Question 4: EPC × Occupancy Interaction

### Create EPC Group

```{r rq4-setup}

# High EPC = A or B
# Low EPC = C to F

data_clean$high_epc <- ifelse(
  data_clean$epc_label %in% c("A", "B"),
  "High",
  "Low"
)

data_clean$high_epc <- factor(
  data_clean$high_epc,
  levels = c("Low", "High")
)

cat("EPC group distribution:\n")

print(
  table(data_clean$high_epc)
)
```

### Additive Model

```{r rq4-additive}

model_additive <- lm(
  annual_kwh ~
    sq_meters +
    high_epc +
    occupancy_count +
    income_euro +
    dist_to_brussels,
  data = data_clean
)
```

### Interaction Model

```{r rq4-interaction}

model_interaction <- lm(
  annual_kwh ~
    sq_meters +
    high_epc * occupancy_count +
    income_euro +
    dist_to_brussels,
  data = data_clean
)

summary(model_interaction)
```

### Nested Model F-Test

```{r rq4-ftest}

anova(
  model_additive,
  model_interaction
)
```

### Interaction Interpretation

```{r rq4-interpretation}

int_coef <- coef(model_interaction)[
  "high_epcHigh:occupancy_count"
]

int_ci <- confint(model_interaction)[
  "high_epcHigh:occupancy_count",
]

int_p <- summary(model_interaction)$coefficients[
  "high_epcHigh:occupancy_count",
  "Pr(>|t|)"
]

cat(
  "Interaction coefficient:",
  round(int_coef, 1),
  "\n"
)

cat(
  "95% CI: [",
  round(int_ci[1], 1),
  ",",
  round(int_ci[2], 1),
  "]\n"
)

cat(
  "p-value:",
  format.pval(
    int_p,
    digits = 3
  ),
  "\n"
)

occ_slope_low <-
  coef(model_interaction)["occupancy_count"]

occ_slope_high <-
  occ_slope_low + int_coef

cat(
  "\nOccupancy slope - Low EPC homes:",
  round(occ_slope_low, 1),
  "kWh per resident\n"
)

cat(
  "Occupancy slope - High EPC homes:",
  round(occ_slope_high, 1),
  "kWh per resident\n"
)

cat(
  "Difference in occupancy slopes:",
  round(abs(int_coef), 1),
  "kWh per resident\n"
)
```

---

## 16. Interaction Plot

```{r interaction-plot, fig.width=7, fig.height=5}

pred_data <- expand.grid(

  occupancy_count = 1:7,

  high_epc = factor(
    c("Low", "High"),
    levels = c("Low", "High")
  ),

  sq_meters =
    mean(
      data_clean$sq_meters,
      na.rm = TRUE
    ),

  income_euro =
    mean(
      data_clean$income_euro,
      na.rm = TRUE
    ),

  dist_to_brussels =
    mean(
      data_clean$dist_to_brussels,
      na.rm = TRUE
    )
)

ci_pred <- predict(
  model_interaction,
  newdata = pred_data,
  interval = "confidence"
)

pred_data$fit <-
  ci_pred[, "fit"]

pred_data$lower <-
  ci_pred[, "lwr"]

pred_data$upper <-
  ci_pred[, "upr"]

interaction_plot <- ggplot(
  pred_data,
  aes(
    x = occupancy_count,
    y = fit,
    colour = high_epc,
    fill = high_epc
  )
) +

  geom_ribbon(
    aes(
      ymin = lower,
      ymax = upper
    ),
    alpha = 0.15,
    colour = NA
  ) +

  geom_line(
    linewidth = 1.1
  ) +

  geom_point(
    size = 2.5
  ) +

  scale_x_continuous(
    breaks = 1:7
  ) +

  labs(
    title =
      "Interaction between EPC group and household occupancy",
    subtitle =
      "Other predictors held at sample means; shaded bands represent 95% confidence intervals",
    x =
      "Number of occupants",
    y =
      "Predicted annual electricity consumption (kWh)",
    colour =
      "EPC Group",
    fill =
      "EPC Group"
  ) +

  theme_minimal(
    base_size = 12
  ) +

  theme(
    legend.position = "bottom"
  )

interaction_plot
```

### Save Interaction Plot

```{r save-interaction-plot}

ggsave(
  "../figures/interaction_plot.png",
  plot = interaction_plot,
  width = 7,
  height = 5,
  dpi = 300
)

cat(
  "Saved: figures/interaction_plot.png\n"
)
```

---

## 17. Session Information

```{r session-info}

sessionInfo()
```

---

*Portfolio version | Project Learning from Data | Hasselt University*
